# A Very Short Introduction to R

```{r global_options_intro, include=FALSE}
# This is a way to set options for all code chunks at once
# Note that you can also dynamically control options by setting them to a value
# DPI setting increased for Word output, cairo is anti aliasing
knitr::opts_chunk$set(eval = FALSE)
```

## Introduction

------------------------------------------------------------------------

Welcome to the first week of Part I - Introduction to R!

The aim of this practical is to introduce you to the R programming language as well as RStudio and some of its functionalities. You will also practice using some basic functions and operations.

The contents of this practical are based on materials developed by Kyle Lang (<https://github.com/kylelang/Introduction-to-R>) as well as Kasper Welbers, Wouter van Atteveldt, Philipp Masur, and Paulina Pankowska (<https://github.com/ccs-amsterdam/r-course-material>).

### What is R and why should you learn it?

R is an open-source statistical software language, that is currently among the most popular languages for data science. In comparison to other popular software packages in social scientific research, such as SPSS and Stata, R has several notable advantages:

-   R is a programming language, which makes it much more versatile. While R focuses on statistical analysis at heart, it facilitates a wide-range of features, and virtually any tool for data science can be implemented.

-   The range of things you can do with R is constantly being updated. R is open-source, meaning that anyone can contribute to its development. In particular, people can develop new *packages*, that can easily and safely be installed from within R with a single command. Since many scholars and industry professionals use R, it is likely that any cutting-edge and bleeding-edge techniques that you are interested in are already available. You can think of it as an app-store for all your data-science needs!

-   R is free. While for students this is not yet a big deal due to free or cheap student and university licences, this can be a big plus in the commercial sector. Especially for small businesses and free-lancers.

## Using RStudio

Once you have installed R and RStudio, you can start by launching RStudio. If everything was installed correctly, RStudio will automatically launch R as well. The first time you open RStudio, you will likely see three separate windows. The first thing you want to do is open an R Script to work in. To do so, go to the toolbar and select File -\> New File -\> R Script.

You will now see four windows split evenly over the four corners of your screen:

-   In the **top-left** you have the text editor for the file that you are working in. This will most of the time be an R script or RMarkdown file.

-   In the **top-right** you can see the data and values that you are currently working with (environment) or view your history of input.

-   In the **bottom-left** you have the console, which is where you can enter and run code, and view the output. If you run code from your R script, it will also be executed in this console.

-   In the **bottom-right** you can browse through files on your computer, view help for functions, or view visualizations.

While you can directly enter code into your console (bottom-left), you should, generally, always work with R scripts (top-left). This allows you to keep track of what you are doing and save every step. There are a few exceptions, which we will cover later.

*Note*: You can also view your R script in a separate window by clicking on the icon with the small upward-pointing arrow in the top-left corner of the script editor. This is similar to the setup in, say, SPSS or Stata, and is especially useful if you have multiple monitors or a large screen.

## Running code from the R script

Copy and paste the following example code into your R Script. For now, don't bother understanding the syntax itself. Just focus on running it.

```{r, eval=F}
3 + 3
2 * 5
(6 / 2) * 3
"some text"
"some more text"
sum(1,2,3,4,5)   ## using a function
```

You can **run** parts of the code in an R script by pressing Ctrl + Enter (on mac this is command + Enter). This can be done in two ways:

-   If you select a piece of text (so that it is highlighted) you can press Ctrl + Enter to run the selection. For example, select the first three lines (the three mathematical operations) and press Ctrl + Enter.

-   If you haven't made a selection, but your text cursor is in the editor, you can press Ctrl + Enter to run the line where the cursor is at. This will also move the cursor to the next line, so you can *walk* through the code from top to bottom, running each line. Try starting on the first line, and pressing Ctrl + Enter six times, to run each line separately.

## Using objects: Assigning values to names

To do anything useful, we need to create objects that hold data.

When running the example code, you saw that R automatically **evaluates** expressions. The calculation 3+3 evaluates to 6, and 2\*5 evaluates to 10. You also saw that the **function** *sum(1,2,3,4,5)* evaluates to 15 (the sum of the numbers). For now, one more thing that you need to know about the R syntax is how values can be **assigned** to objects via the 'assignment' operator.

In plain terms, **assignment** is how you make R remember things by assigning them a name. This works the same way for all sorts of values, from single numbers to entire datasets. You can choose whether you prefer the equal sign (=) or the arrow (\<-) for assignment.

```{r}
x = 2
x2 <- 33.33
y <- 4

"Bob" -> z
```

Here we have, among other things, remembered the number 2 as **x** and the text "some text" as **y**. If you are working in RStudio (which you should), you can now also see these names and values in the topright window, under the "Environment" tab.

Evaluating an object name without assignment prints the value of that object (i.e., use the names to retrieve the values)

```{r}
y
x
z
```

We can also use these values in new commands.

```{r}
x * 5
```

*Note*

The assigned values can also be named objects

```{r}
w <- y
w
```

Object names must begin with a letter

```{r}
my1X <- pi
my1X

# 1X <- pi # this will not work :(
```

### PRACTICE PROBLEM 1

Create an object named `test` and assign the sum of 10 and 15 to it.

```{r}
# SOLUTION
```

## Comments

Comments are a useful way to annotate or document your code, so you and others can more easily understand it and know what is going on.

The comment character in R is '\#'; each commented line must be preceded by a '\#' symbol as there are no block comments in R. Comments are not evaluated when you run your code

## Installing Packages

Packages are collections of R functions (which we'll talk about more later. For now it is enough to know that they are a set of statements organized together to perform a specific tasks; they are similar to commands in Stata or SPSS), data, and compiled code in a well-defined format, created to add specific functionality. There are 10,000+ user contributed packages and growing.

There are a set of standard (or base) packages which are considered part of the R source code and automatically available as part of your R installation. Base packages contain the basic functions that allow R to work, and enable standard statistical and graphical functions on datasets. Throughout our course you will need to install various packages, for example, for web scraping or network analysis.

The standard way to install packages is using the following command:

```{r}
install.packages("psych")
```

This command will install the package 'psych' from the Comprehensive R Archive Network (CRAN). CRAN is a network of servers around the world that store identical, up-to-date, versions of code and documentation for R. There are also ways to install packages from other places, which we don't cover here for simplicity.

### PRACTICE PROBLEM 2

Use `install.packages()` to install the following packages in the default location (i.e., don't specify anything for the 'lib' argument): - ggplot2 - dplyr - haven

```{r}
# SOLUTION 
```

Much like any other software, we only need to install a package once on our computer (unless we update R), but we need to load the package each time we open R and want to use it.

```{r}
library(psych)
```

While we recommended that you generally run commands from scripts rather than from the console, "install.packages()" may be an exception: as you typically need to run this only once per package, it is often more practical to run it directly from the console.

## The working directory

Every R session is associated with a 'working directory'. The working directory is the directory wherein R will root its navigation when reading or writing data objects to or from disk.

Find the current working directory

```{r}
getwd()
```

Change the current working directory

```{r}
setwd("/Users/corte101/Teaching") # obviously this will not work for you, unless you are me!
getwd()
```

*Note:* Although opinions on this differ, this may be another exception to the rule that you should run commands from scripts rather than from the console. Setting the working directory is somwthing you may want to do outside your script, as it is a setting that is specific to your computer and your file structure, and you may not want your script to depend on that to work correctly.

### PRACTICE PROBLEM 3

Use the `setwd()` to change your working directory to the directory in which this script is saved.

```{r}
# SOLUTION 
```

## Interacting with the Environment

The 'environment' is a loosely organized set of all the objects that R currently has stored in working memory.

We can check the contents of the current environment using a command:

```{r}
ls()
```

...but you can also see the content of your environment in the top-right window of RStudio, under the "Environment" tab.

And remove an object from the environment

```{r}
rm(x)
ls()
```

Or completely clear the enviroment

```{r}
rm(list = ls(all = TRUE))
ls()
```

### PRACTICE PROBLEM 4

Create two objects and assign any value to them. Then use `rm()` to remove one of the objects.

```{r}
# SOLUTION
```

## Back to writing code

### Mathematical Operators

Arithmetic

```{r}
x<-2
y<-4

y + x
y - x
y * x
y / x
```

Powers

```{r}
y^2
y^3
```

Roots

```{r}
sqrt(y)
```

#### PRACTICE PROBLEM 5

(a) Create an object called 'age' that takes the value of your age in whole years.
(b) Use the 'age' object you created in (a) to create a second object called 'weeks' that takes the value of your age in whole weeks (assume 52 weeks in each year and disregard partial years).

```{r}
# SOLUTION
```

### Logical Comparisons

```{r}
y <- 5
x <- 7
w <- 5
```

Check equality

```{r}
y == x
y == w
```

Check relative size

```{r}
y > x  # greater than
y >= x # greater than or equal to
y < x  # less than
y <= x # less than or equal to

y > w
y >= w
y < w
y <= w
```

We can negate any logical condition by prepending the '!' character

```{r}
y > x
!y > x

y == w
y != w
```

We can create more complex logical conditions with the AND and OR operators: '&' and '\|'

```{r}
y == w & y < x
y == w & y > x
y == w | y > x
```

#### PRACTICE PROBLEM 6

Use a single line of code to generate a logical value (i.e., TRUE/FALSE) indicating if the value of the `weeks` object you created above is larger than the `age` object you created.

```{r}
# SOLUTION
```

## Data structures

In SPSS or Stata, data is organized in a rectangular data frame, with cells arranged in rows and columns.

Typically, the rows then represent cases (e.g., respondents, participants, countries) and columns represent variables (e.g., age, gender, education level, GDP).

For most analyses, this is also the recommended data format in R, using the *data.frame* structure. However, an important difference is that in R it is possible, and often useful, to combine different formats.

Also, to understand how a *data.frame* in R works, it is useful to understand that a *data.frame* is a collection of *vectors*, and thus it is useful to first understand how *vectors* work.

Here we will first briefly discuss *vectors*, and then quickly move on to *data.frames*. In addition, there are other common data structures, such as the matrix and list, which we will not discuss here.

Different packages can also provide new classes for organizing and manipulating data.

### Vectors

The concept of a vector might be confusing from a social science background, because we rarely use the term in the context of statistics (well, not consciously at least). We won't address why R calls them vectors and how this relates to vector algebra, but only how you most commonly use them.

A vector in R is a **sequence** of **one or more values** of the **same data type**. From a social science background, it is very similar to what we often call a **variable**.

You can declare a vector in R with `c(...)`, where between the parentheses you enter the elements, separated with commas.

The number of elements is called the length of the vector. A vector can have any of the data types discussed above (numeric, character, factor, logical, Date).

```{r}
v1 <- c(1, 2, 10, 15)    ## a numeric vector of length 4
v2 <- c("a", "b", "b")   ## a character vector of length 3
v3 <- 1:10               ## a numeric vector of length 10 with the values 1 to 10 
```

If you combine data types in the same vector, R will generally use the broadest data type for the entire vector. For example, if we combine both types in a vector, R will convert the numerical values to character values.

```{r}
c(1, 2, "c")            ## becomes a character vector of length 3
```

#### Selecting elements

There are two common ways to select a specific element or a range of elements from a vector. One is to give the indices (positions) of the elements in square brackets after the vector name. Note that the indices themselves are given as a numeric vector.

```{r}
x <- c('a','b','c','d','e','f','g')  
x[5]            ## select the fifth element
x[c(1,3)]       ## select the first and third elements
x[2:5]          ## select elements two to five
```

If you select with indices, the specific order of the indices is used, and you can also repeat indices. This can for instance be used to sort data.

```{r}
x[5:1]          ## select elements in positions 5 to 1
x[c(5,5,5)]     ## select the element in position 5 multiple times
```

You can also use negative indices to select everything except the specified elements.

```{r}
x[-5]            ## select every element except the fifth
x[-c(1,3)]       ## select every element other than the first and third
```

The second way to select values is to use a **logical vector**, which we don't cover here.

### Data frames

A *data frame* is essentially a collection of vectors with the same length, tied together as columns. This is equivalent to a data matrix that you would use in a program such as SPSS. To create the data frame*, we use `data.frame()`. We enter the vectors in the format: `column_name = vector`. Here we create a* data.frame\* for data from a fictional experiment.

```{r, results=T}
d <- data.frame(id =        1:10,
                condition = c('E', 'E', 'C', 'C', 'C', 'E', 'E', 'E', 'C', 'C'),
                gender =    c('M', 'M', 'F', 'M', 'F', 'F', 'F', 'M', 'M', 'F'),
                age =       c( 17,  19,  22,  18,  16,  21,  18,  17,  26,  18),
                score_t1 =  c(8.0, 6.0, 7.5, 6.8, 8.0, 6.4, 6.0, 3.2, 7.3, 6.8),
                score_t2 =  c(8.3, 6.4, 7.7, 6.3, 7.5, 6.4, 6.2, 3.6, 7.0, 6.5))
d
```

Now, the data structure clearly implies that there is a relation between the elements in the *column vectors*.

In other words, that each row represents a *case*. In our example, these cases are participants, and the columns represent:

-   the participant **id**.
-   the experimental **condition** (E = experimental condition, C = control group)
-   demographic variables: **gender** and **age**.
-   test scores before and after the experimental condition: **score_t1** and **score_t2**

### Selecting rows, columns and elements in data frames

Since data.frames have both rows and columns, we need to use both to select data. Similar to selection in vectors, we use the square brackets. The difference is that for data.frames the square brackets have two parts, separated by a comma. Assuming our data.frame is called `d`, we can select with:

| syntax | meaning                                  |
|--------|------------------------------------------|
| d[i,j] | select rows (i) and columns (j)          |
| d[i, ] | select only rows (i) and use all columns |
| d[ ,j] | select only columns (j) and use all rows |

Selection for rows (i) and columns (j) works identical to selection in vectors. You can use either a numeric vector with indices, or a logical vector. Accordingly, you can also use comparisons.

In addition, there are two special ways to select columns. One is that j can be a character vector with column names. The other uses the dollar sign (\$).

| syntax             | meaning                                         |
|--------------------|-------------------------------------------------|
| d[ ,c("c1", "c2")] | select the columns with the names "c1" and "c2" |
| d\$id              | select the column named id                      |

#### Selecting columns in data frames

Let's put this to practice, starting with columns:

```{r}
## selecting a single column returns a vector
d[,1]             ## select the first column by index 
d[,"id"]          ## select the id column by name
d$id              ## select the id column using the dollar sign
## selecting multiple columns returns a data.frame
d[,1:2]           ## select the first two columns by indices
d[,c("id","age")] ## select the "id" and "age" columns by name
d[,-1]            ## select every column except for the first  
```

#### Selecting rows in data frames

Selecting rows is practically identical to selecting elements from vectors, and it conveniently returns a data.frame with all columns and their matched positions intact.

```{r}
d[1:5,]    ## select first 5 rows
```

A very useful additional trick is that you can use all the columns to make comparisons. For example, we can use the gender column to look up all elements for which the value is "M" (male), and use this to select rows.

```{r}
d[d$gender == "M", ]       
```

You can combine this with the logical operators to make a selection using multiple columns.

```{r}
d[d$gender == "F" & d$age == 21, ]    ## 21 year old female participant(s)
d[d$score_t1 < d$score_t2,]           ## participants that scored higher after the condition
```

#### Selecting rows and columns in data frames

We can combine row and column selection. This works just like you'd expect it to, so there's little to add here. Do note, however, that you can combine the different selection methods.

```{r}
d[d$gender == "F", "score_t1"]    ## get the score_t1 column for female participants
d[d$gender == "F",]$score_t1      ## identical, but first subset data.frame, then select column
d$score_t1[d$gender == "F"]       ## identical, but first select column vector, then subset vector
```

### PRACTICE PROBLEM 7

*Subsetting, adding and modifying data*

With the selection techniques you already learned how to create a subset of the data. Try to subset the data so that only participants in the condition "E" are included. Assign this subset to a new name. Create a second subset in which only the variables *id*, *gender*, and *age* are included. Assign this subset to a different name.

```{r}
# SOLUTION
```

## Getting data into R

There are a number of ways to read data into R. One option is to use the `readr` package that comes with the `tidyverse` distribution, `read_csv()` reads comma delimited files.

Download the file "flightdata.csv" from [here](https://raw.githubusercontent.com/rensec/srda/main/flightdata.csv) and store it in your project folder (same folder that you set as your working directory at the beg. of the practical). Most likely, the csv file will open in your browser; just choose "Save as..." (or equivalent) from your browser's "File" menu so save the file to disk. This file contains a sample from the "flights" dataset from the `nycflights13` package. This contains airline data for all flight departing from NYC in 2013.

Read the `flightdata.csv` file into R with the `readr` package\` using the code below

```{r message=FALSE, warning=FALSE}
# install.packages("readr")
library(readr)

flight_data <- read_csv("flightdata.csv") # Imports the data
flight_data # View the data
```

To get other types of data into R the tidyverse packages listed below are recommended.

-   `haven` reads SPSS, Stata, and SAS files
-   `readxl` reads excel files (.xls and .xlsx)

### Summarising the data

Again, there are different functions to summarise data, but `summary()` that is available in base R works well too.

Applying `summary()` to the data

```{r}
summary(flight_data)

# `summary()` returns various summary statistics for every numeric variable including minimum and maximum, quartiles, mean and median. Notice that for character type variables, this returns the length of the variable.
```

## Saving data

Of course, you can also save a dataset in our computer in a variety of formats (e.g., .RData, .csv...). When saving, it is always important to understand *where* you are saving the file. Each R session is connected to a particular folder on your computer. you can check where you are by using `getwd()` which you were introduced to earlier. This resulting path is where the data will be stored.

```{r, eval = F}
# Where are we currently on our computer?
getwd()
# Save and load an RData file
save(d, file = "test_data.RData")   ## file = file and/or connection to write to
load(file = "test_data.RData")

# Save and load a csv file
write_csv(d, file = "test_data.csv")
d2 <- read_csv(file = "test_data.csv", )
d2
```

## Functions

Where data types and structures concern how data is represented in R, functions are the tools that you use to read, create, manage, manipulate, analyze and visualize data.

A function is a set of statements organized to perform a specific task. It is an object that accepts a certain number of arguments and then returns the result of the task performed on these arguments.

R has a large number of built-in functions and you can also create your own functions. For now, we will focus on the former. Simple examples of built-in functions are `seq()`, `mean()`, `min()`, `max()`, and `sum()`.

While there are many correct and formal ways to define what functions are, for the sake of simplicity we will focus on an informal description of how you can think of functions in R:

-   A function is used in the following way: `output <- function_name(argument1, argument2, ...)`
    -   **function_name** is a name to indicate which function you want to use. It is followed by parentheses.
    -   **arguments** are the input of the function, and are inserted within the parentheses. Arguments can be any R object, such as numbers, strings, vectors and data.frames. Multiple arguments can be given, separated by commas.
    -   **output** is anything that is returned by the function, such as vectors, data.frames or the results of a statistical analysis. Some functions do not have output, but produce a visualization or write data to disk.
-   The purpose of a function is to make it easy to perform a (large) set of (complex) operations. This is crucial, because
    -   It makes code easier to understand. You don't need to see the operations, just the name of the function that performs them.
    -   You don't need to understand the operations, just how to use the function.

For example, say that you need to calculate the square root of a number. This is a very common thing to do in statistical analysis, but it actually requires a quite complicated set of operations to perform. This is when you want to use a function, in this case the `sqrt` (square root) function.

```{r}
sqrt(5)
```

In this example, the function name is `sqrt`. The input is the single argument `5`. If you run this code, it produces the output `2.236068`. Currently, R will simply print this output in your Console, but as you learned before, we can assign this output to a name.

```{r}
square_root <- sqrt(5)
```

This simple process of input -\> function -\> output is essentially how you work with R most of the times. You have data in some form. You provide this data as input to a function, and R generates output. You can assign the output to a name to use it in the next steps, or the output is a table with results or a visualization that you want to interpret.

### Using functions

Above you saw the simple function `sqrt()`, that given a single number as input returned a single number as output. Functions can also have multiple arguments as input. Consider this function here:

```{r}
mean(x = c(1, 2, NA, 1, 2, 5), na.rm = TRUE)
```

This function, with the name `mean`, is given several arguments here: `x` and `na.rm`. Given this input, many operations are performed behind the scenes to create the arithmetic mean of the vector.

By now we hope you have realized just how broad the use of functions is. The *R syntax* for performing basic mathematical operations such as `sqrt()` is essentially the same as the syntax for creating a mean, performing advances statistical analysis or creating powerful visualizations. Accordingly, if you understand this syntax, you can do almost anything in R.

### PRACTICE PROBLEM 8

Perform the following operations: 1. Create a sequence of numbers from 12 to 24. HINT: look up ?s eq 2. Sum the numbers from 20 to 60. 3. Find the mean of the numbers from 25 to 80.

```{r}
# SOLUTION
```

The many R packages that you can install are mostly just collections of functions. We realize that it is a lot and might be a bit overwhelming. So we will now show you how to use the R help option which allows you to view and interpret any function documentation page and learn how to use it.

## Getting help: viewing and interpreting function documentation

You can access the documentation of a function by typing a question mark in front of the function name, and running the line of code. Let's do this to view the documentation of the `sqrt()` function

```{r}
?sqrt
```

Alternatively, you can also use the 'help' function

```{r}
help(sqrt)
help("sqrt")
```

If you run this in RStudio, the help page will pop-up in the bottom-right corner, under the *Help* tab page. Sometimes, if the name of a documentation page is used in multiple packages, you will first receive a list of these packages from which you will have to select the page.

For the `sqrt()` function, the help page has the **title** "Miscellaneous Mathematical Functions". Just below the title, you see the **Description**, in which the author of a function briefly describes what the function is for.

Here we see that there are two functions that are grouped under "Miscellaneous Mathematical Functions", the `abs()` function for computing the absolute value of a number `x`, and the `sqrt()` function for the square root.

Under description, the **Usage** is shown. This is simply the name of the function or functions, and the possible arguments that you can use. Here the Usage is extremely simple: both functions only take one argument named `x`. In a minute, we'll discuss functions with multiple arguments.

Below usage, the **Arguments** section explains how to use each argument. Here, the only argument is `x`, and it is explained that x is "a numeric or complex vector or array". For now, let's focus only on the case of a numeric vector. It appears that in addition to giving a single value like above (recall that in R this is actually a vector of length 1) we can give a vector with multiple numbers.

```{r}
sqrt(c(1,2,3,4,5))
```

There are more parts to the documentation that we'll ignore for now. Notable parts to look into for yourself are **Details**, that provides more information, and the **Examples** section at the very bottom, which is a great starting point to see a function in action.

*Note* Non-letter characters need to be quoted

```{r}
# ?/ will not work
# help(/) will not work 

?"/" #  will work
help("/") #  will work
```

*Note II* If a package is not loaded, you need to specify the namespace

```{r}
install.packages("mice", repos = "http://cran.us.r-project.org")
# ?quickpred  will not work
?mice::quickpred
help(quickpred, package = "mice")
```

*Note III* You can also open an interactive web-based help page

```{r}
help.start()
```

### PRACTICE PROBLEM 9

(a) Access the help file for the vector() function.
(b) How many arguments does the vector() function take?

```{r}
# SOLUTION 
```
